A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-06-01 bioc-workflow.Rmd
Updates to ML section

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
[@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT),
an improved method relying on improved sub-cellular fractionation and
more acurate quantitation, leading to more reliable classification of
protein localisation across the whole sub-cellular space. The method
uses an elaborate sub-cellular fractionation scheme, enabled by the
use of Tandem Mass Tag (TMT) [Thompson:2003] 10-plex and application
of the MS data acquisition technique named synchronous precursor
selection MS$^3$ (SPS-MS$^3$) [@McAlister:2014], for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was
to target low density fractions and the second and third were to
emphasis separation of the denser organelles.  The intersect of
replicates 1 and 2 was treated as a 20-plex dataset for the analysis
discussed in the manuscript [@hyper] as it has been shown that
combining replicates across from different gradients can increase
spatial resolution [@trotter]. The combination of replicates resulted
in 5032 proteins common in both experiments.

These, as well as many other data are directly available as properly
structured and annotated computational object from the 
`r Biocexptpkg("pRolocdata")` experiment package. In this workflow, we
will start with a description of how to generate these ad hoc objects
starting from any arbitrary spreadsheets, as produced by many popular
third-party applications.

While we focus here on a LOPIT-type dataset, these analyses are
relevant for any quantitative spatial proteomics data, irrespective of
the fractionation or quantitation (i.e. labeled or label-free)
methods.

@@

To make use of the full functionality of the `r Biocpkg("pRoloc")`
software one needs to import their data into R and prepare them as an
`MSnSet`. The `MSnSet` is a dedicated data structure for the efficient
manipulation and processing of mass spectrometry and proteomics data
in R. Figure 1 illustrates a simplified view of the `MSnSet`
structure; there exists 3 key sub-parts (termed slots) to such a data
object: (1) the `exprs` slot for storing the quantitation data, (2)
the `fData` slot for storing the feature meta-data, and finally (3)
the `pData` slot for storing the sample meta-data.

![Simplified representation of the `MSnSet` data structure (reproduced with permission from the `r Biocpkg("MSnbase")` vignette)](./Figures/msnset.png)

There are a number of ways to import quantitation data and create an
`MSnSet` instance and all methods are described in the `r
Biocpkg("MSnbase")`
[input/output capabilities vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf). One
suggested simple method is to use the function `readMSnSet2` in 
`r Biocpkg("MSnbase")`. The function takes a single spreadsheet as input
and extracts the columns containing the quantitation data, as
identified by the argument `ecol`, to create the expression data,
while the other columns in the spreadsheet are appended to the feature
meta-data slot.  By example, in the code chunk below we read in the
`csv` spreadsheet containing the quantitation data from the intersect
of replicates 1 and 2 of the mouse map [@hyper], using the
`readMSnSet2` function. The data is as available online with the
manuscript (see tab 2 of the `xlsx` supplementary data set 1 in
[@hyper], which should be exported as a text-based spreadsheet) and
also as a `csv` in the Bioconductor `r Biocexptpkg("pRolocdata")` data
package.

To use the `readMSnSet2` function, as a minimum one must specify the
file path to the data and which columns of the spreadsheet contain
quantitation data. The `getEcols` function exists to help users
identify which columns of the spreadsheet contain the
quantitation data.

We first locate the the spreadsheet of E14TG2a data distributed with
the `r Biocexptpkg("pRolocdata")` data using `system.file`, that
locates the `extdata` in the `r Biocexptpkg("pRolocdata")` package on
the hard drive and `dir`, that displays the full path to the file
matching the `hyperLOPIT-SIData-ms3-rep12-intersect.csv` pattern,
which corresponds the the file of interest. In the last line, we print
the filename (not the full path, which will vary from computer to
computer).

```{r getFilename}
library("MSnbase")
extdatadir <- system.file("extdata", package = "pRolocdata")
csvfile <- dir(extdatadir, full.names = TRUE,
          pattern = "hyperLOPIT-SIData-ms3-rep12-intersect.csv")
basename(csvfile)
```

Note that the file is compressed (as indicated by the `gz`, for
`gzip`, extension), and will be decompressed on-the-fly when read into
R later on.

The spreadsheet that was deposited by the authors contains two
headers, with the second header containing information about where the
quantitation data is stored. We can display the names of the second
header by calling the `getEcols` function with the argument `n = 2`
(the default value is `n = 1`), to specify that we wish to display the
column names of the second line.

```{r getEcols}
getEcols(csvfile, split = ",", n = 2)
```

It is now easy for one to
identify that the quantitation data is located in columns 8 to 27. It
is also possible to pass the optional argument `fnames` to indicate
which column to use as the labels by which to identify each protein in
the sample. Here, we use `fnames = 1` to use the Uniprot identifiers
contained in the first column of the spreadsheet.

```{r readMSnSet2}
lopit2016 <- readMSnSet2(csvfile, ecol = c(8:27), fnames = 1, skip = 1, 
                         stringsAsFactors = FALSE)

@@

contains 5032 proteins common across the 2 biological replicates for
the respective 2 x 10-plex reporter tags, along with associated
feature meta-data such as protein markers, protein description, number
of quantified peptides etc.

@@

There are several classification algorithms available in `pRoloc`, which are documented in the dedicated [`pRoloc` Machine learning techniques vignette](http://bioconductor.org/packages/release/bioc/vignettes/pRoloc/inst/doc/pRoloc-ml.pdf). We find the general tendancy to be that it is not the choice of classifier but the proper optimisation of the algorithms parameters that limit the classification results. Before employing a classification algorithm and generating a model on the training data with which to classify our set of unknown residents, one must find the optimal parameters for the algorithm of choice. For example, in the code chunk below we employ the use of a Support Vector Machine (SVM) to learn a classifier on the labelled training data. In the example below, the training data is found in the `featureData` slot in the column called `SVM.marker.set`.

@@

```{r loadpdRes, include=FALSE}
pdRes <- lopit2016
load("Data/fusionparams.rda")
```
```{r optimise, eval=FALSE}
w <- table(fData(pdRes)[, "SVM.marker.set"])
w <- 1/w[names(w) != "unknown"]

params <- svmOptimisation(pdRes, fcol = "SVM.marker.set",
                          times = 100, xval = 5,
                          class.weights = w,
                          verbose = TRUE)
```
```{r visualiseOpt}
plot(params)
levelPlot(params)
(best <- getParams(params))
```
```{r classify, eval=TRUE}
svmRes <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "SVM.marker.set")
```

>>>>

As a use-case we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
[@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT),
an improved method relying on better sub-cellular fractionation and
more acurate quantitation, leading to more reliable classification of
protein localisation across the whole sub-cellular space. The method
uses an elaborate sub-cellular fractionation scheme, enabled by the
use of Tandem Mass Tag (TMT) [Thompson:2003] 10-plex and application
of the MS data acquisition technique named synchronous precursor
selection MS$^3$ (SPS)-MS$^3$ [@McAlister:2014], for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was
to target low density fractions and the second and third were to
emphasis separation of the denser organelles. The intersect of
replicates 1 and 2 was treated as a 20-plex dataset for the analysis
discussed in the manuscript [@hyper] as it has been shown that
combining replicates across from different gradients can increase
spatial resolution [@trotter]. The combination of replicates resulted
in 5032 proteins common in both experiments.

* Need to say that can be label-free or any other type of
  quantitation, and, possibly, that features can be a peptide level.

* I would like to show code for the below - maybe move to later
  sections. See *Data processing* section.

@@

To make use of the full functionality of the `r Biocpkg("pRoloc")` software one is required to import their data into R as a `MSnSet` instance. The `MSnSet` is a dedicated data structure for the efficient manipulation and processing of mass spectrometry and proteomics data in R. Figure 1 illustrates a simplified view of the `MSnSet` structure; there exists 3 key slots (1) the `exprs` slot for storing the quantitation data, (2) the `fData` slot for storing the feature meta-data, and finally (3) the `pData` slot for storing the sample meta-data. 
![**Figure 1.** Simplified representation of the `MSnSet` data structure (reproduced with permission from the `r Biocpkg("MSnbase")` vignette)](./Figures/msnset.png)

There are a number of ways to import quantitation data and create a `MSnSet` instance and all methods are described in the `r Biocpkg("MSnbase")` [input/output capabilities vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf). One suggested simple method is to use the function `readMSnSet2` in `r Biocpkg("MSnbase")`. The function takes a single spreadsheet as input and extracts the columns containing the quantitation data, as identified by the argument `ecol`, to create the expression data, while the other columns in the spreadsheet are appended to the feature meta-data slot. 
By example, in the code chunk below we read in the `csv` spreadsheet containing the quantitation data from the intersect of replicates 1 and 2 of the mouse map [@hyper], using the `readMSnSet2` function. The data is as available online with the manuscript (see tab 2 of the `xlsx` supplementary data set 1 in [@hyper]) and also as both a `csv` and `MSnSet` object in the Bioconductor `r Biocexptpkg("pRolocdata")` data package.

To use the `readMSnSet2` function, as a minimum one must specify the file path to the data and which columns of the spreadsheet contain quantitation data. The `getEcols` function exists to help users identify which columns of the spreadsheet that contain the quantitation data. The spreadsheet of E14TG2a data; "hyperLOPIT-SIData-ms3-rep12-intersect.csv", contains 5032 proteins common across the 2 biological replicates for the respective 2 x 10-plex reporter tags, along with associated feature meta-data such as protein markers, protein description, number of quantified peptides etc. The spreadsheet contains two headers, with the second header containing information about where the quantitation data is stored. We can display the names of the second header by calling the `getEcols` function with the argument `n = 2`, to specify that we wish to display the column names of the second header. It is now easy for one to identify that the quantitation data is located in columns 8 to 27. It is also possible to pass the optional argument `fnames` to indicate which column to use as the labels by which to identify each protein in the sample. Here, we use `fnames = 1` to use the Uniprot identifiers contained in the first column of the spreadsheet.

```{r makeMSnSet, message=FALSE, warning=FALSE}
library("MSnbase")

f0 <- dir(system.file("extdata", package = "pRolocdata"), full.names = TRUE, 
          pattern = "hyperLOPIT-SIData-ms3-rep12-intersect.csv")

getEcols(f0, split = ",", n = 2)

lopit2016 <- readMSnSet2(f0, ecol = c(8:27), fnames = 1, skip = 1, 
                         stringsAsFactors = FALSE)

@@

@@

There are several classification algorithms available in `pRoloc`, which are documented in the dedicated [`pRoloc` machine learning techniques vignette](http://bioconductor.org/packages/release/bioc/vignettes/pRoloc/inst/doc/pRoloc-ml.pdf). We find the general tendancy to be that it is not the choice of classifier, but the optimisation of the algorithmic parameters, that limits classification accuracy. Before employing any classification algorithm and generating a model on the training data with which to classify our set of unknown residents, one must find the optimal parameters for the algorithm of choice. In the code chunk below we employ the use of a Support Vector Machine (SVM) to learn a classifier on the labelled training data. The training data is found in the `featureData` slot in the column called `SVM.marker.set`. As previously mentioned one first needs to train the classifiers parameters before an algorithm can be used to predict the class labels of the proteins with unknown location. One of the most common ways to optimise the parameters of a classifier is to partition the labelled data in to training and testing subsets. In this framework parameters are tested via a grid search using cross-validation on the training partition. The best parameters chosen from the cross-validation stage are then used to build a classifier to predict the class labels of the protein profiles on the test partition. Observed and expected classication results can be compared, and then used to assess how well a given model works by getting an estimate of the classiers ability to achieve a good generalisation i.e. that is given an unknown example predict its class label with high accuracy. In `pRoloc` algorithmic performance is estimated using stratified 80/20 partitioning for the training/testing subsets respectively, in conjuction with five-fold cross-validation in order to optimise the free parameters via a grid search. This procedure is usually repeated 100 times and then the best parameter(s) are selected upon investigation of classifier accuracy, here we use the harmonic mean of precision and recall; the macro F1 score. In the code chunk below we demonstrate how to optimise the free parameters; `sigma` and `cost`, of a classical SVM classifier with a Gaussian kernel using the function `svmOptimisation`. As the number of labelled instances per class varies from organelle to organelle, we can account for class imbalance by setting specific class weights when generating the SVM model. Below the weights, `w` are set to be inversely proportional to the class frequencies.

```{r loadpdRes, include=FALSE}
pdRes <- lopit2016
load("Data/fusionparams.rda")
```
```{r optimise, eval=FALSE}
(w <- table(fData(pdRes)[, "SVM.marker.set"]))
w <- 1/w[names(w) != "unknown"]

## 100 rounds of optimisation with five-fold cross-validation
params <- svmOptimisation(pdRes, fcol = "SVM.marker.set",
                          times = 100, xval = 5,
                          class.weights = w,
                          verbose = FALSE)
```

The output `params` is an object of class `GenRegRes`; a regularisation/optimisation framework container for the optimisation results.

associated macro F1 scores. A high macro F1 score
indicates that the marker proteins in the test dataset are consistently correctly assigned by
17
Spatial proteomics data analysis
the algorithm. Often more than one parameter or set of parameters gives rise to the best
generalisation accuracy. As such it is always important to investigate the model parameters
and critically assess the best choice. The best choice may not be as simple as the parameter
set that gives rise to the highest macro F1 score and one must be careful to avoid overtting
and to choose parameters wisely.

```{r visualiseOpt}
plot(params)
levelPlot(params)
(best <- getParams(params))
```
```{r classify, eval=TRUE}
svmRes <- svmClassification(pdRes, params,
                            class.weights = w,
                            fcol = "SVM.marker.set")
```

@@
